Plasticity and adaptation to high light intensity amplify the advantage of amphistomatous leaves

### 2025-09-15 ms/ms.tex
reformat ms

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\KOMAoption{captions}{tableheading}
\usepackage{hyperref}
\addtokomafont{disposition}{\rmfamily}
\usepackage{siunitx}
\newcommand{\aahigh}{\mathrm{AA}_{2000}}
\newcommand{\aalow}{\mathrm{AA}_{150}}
\newcommand{\aashade}{\mathrm{AA}_{\text{shade}}}
\newcommand{\aasun}{\mathrm{AA}_{\text{sun}}}
\newcommand{\aax}{$\mathrm{AA}$}
\newcommand{\aay}{\mathrm{AA}}
\newcommand{\Aamphi}{$A_{\mathrm{amphi}}$}
\newcommand{\Ahypo}{$A_{\mathrm{hypo}}$}
\newcommand{\agcurve}{$A \textendash g_\text{sw}$}
\newcommand{\aqcurve}{$A \textendash Q$}
\newcommand{\ca}{$C_\mathrm{a}$}
\newcommand{\caequals}[1]{$C_\mathrm{a} = \qty{#1}{\micro\mol\raiseto{-1}\mol}$}
\newcommand{\cabetween}[2]{#1 < $C_\mathrm{a} < \qty{#2}{\micro\mol\raiseto{-1}\mol}$}
\newcommand{\gias}{$g_\text{ias}$}
\newcommand{\gmax}{$g_\text{max}$}
\newcommand{\gmaxratio}{$g_\text{max,ratio}$}
\newcommand{\gmaxshade}{$g_\text{max,shade}$}
\newcommand{\gmaxsun}{$g_\text{max,sun}$}
\newcommand{\gsw}{$g_\text{sw}$}
\newcommand{\loggsw}{$\log(g_\text{sw})$}
\newcommand{\logA}{$\log(A)$}
\newcommand{\ppfd}{$\mathrm{PPFD}$}
\newcommand{\ppfdequals}[1]{$\mathrm{PPFD} = \qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
\newcommand{\ppfdqty}[1]{$\qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
\newcommand{\rh}{$\mathrm{RH}$}
\newcommand{\rhequals}[1]{$\mathrm{RH} = #1\%$}
\newcommand{\tleafequals}[1]{$T_\mathrm{leaf} = \qty{#1}{\degreeCelsius}$}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

@@

\begin{center}
$^1$School of Life Sciences, University of Hawaiʻi; Mānoa, Hawaiʻi, USA\\
$^2$Department of Botany, University of Wisconsin, Madison; Wisconsin, USA\\
\medskip
*Corresponding author: cdmuir@wisc.edu\\
\medskip
ORCID: Christopher D. Muir — \href{https://orcid.org/0000-0003-2555-3878}{0000-0003-2555-3878}\\
\end{center}

@@

@@

Stomata are microscopic pores on the surfaces of leaves and other
photosynthetic organs formed by a pair of guard cells. They are
essential for balancing carbon gained per unit water lost and were an
essential innovation in vascular plants, permitting them to grow tall on
land by enabling access to CO\(_2\) for photosynthetic carbon
assimilation while preventing hydraulic failure in variable environments
(\emph{1}--\emph{3}). Optimal stomatal function depends on both dynamic
changes in aperture on the scale of minutes to hours, as well as static
anatomy determined by developmental plasticity and constitutive genetic
differences (\emph{4}--\emph{6}). Understanding how stomata respond to
environmental change over daily, developmental, and evolutionary time is
important for studying adaptation (\emph{1}, \emph{7}, \emph{8}),
inferring paleoclimate from fossil cuticles (\emph{9}--\emph{11}),
predicting responses to climate change (\emph{12}--\emph{15}), and
improving crops (\emph{16}). Stomatal function contributes to global
carbon and water cycles (\emph{17}) and therefore to predicting future
climate (\emph{18}).

Despite extensive theoretical and empirical progress understanding
stomata function and anatomy from molecular to ecosystem levels, the
adaptive significance of amphistomatous leaves remains an important
unsolved problem in leaf structure-function relationships
(\emph{19}--\emph{26}). Amphistomatous leaves develop abaxial and
adaxial stomata whose aperture can be independently regulated
(\emph{27}--\emph{31}) to control gas exchange through each surface. All
else being equal, simultaneous gas exchange through stomata on both
surfaces increases CO\(_2\) supply to chloroplasts by providing a second
parallel pathway through leaf intercellular airspaces, enhancing
photosynthesis (\emph{20}, \emph{32}). The extent to which amphistomy
increases CO\(_2\) supply depends on resistance to diffusion in
intercellular airspaces. This resistance can be low in thin, porous,
amphistomatous leaves (\emph{28}, \emph{33}), but may be more
substantial in thick, dense, hypostomatous leaves (\emph{34}). We refer
to the intercellular airspace conductance (\gias), the inverse of
resistance. Amphistomatous leaves also lose more water through
evaporation because of a second boundary layer conductance (\emph{35}),
but the additional carbon gain should be enough to offset this cost in
most realistic scenarios (\emph{36}).

The paradoxical fact is that, despite the photosynthetic benefit, most
leaves are not amphistomatous. Many vertically oriented and/or
isobilateral leaves are amphistomatous (\emph{25}). But among
dorsiventral leaves, it is primarily herbaceous plants in open, high
light habitats that tend to have amphistomatous leaves (\emph{22},
\emph{39}--\emph{44}). Most other leaves, except those from aquatic
habitats, are hypostomatous, producing stomata only on the lower,
abaxial surface. Even resupinate leaves develop stomata on the lower,
albeit adaxial surface (\emph{45}), suggesting that leaf orientation
(lower vs.~upper) rather than leaf polarity (abaxial vs.~adaxial) is
causal. Stomatal density ratio, defined as the ratio of upper to total
stomatal density, is a quantitative metric of stomatal distribution. The
covariation between stomatal density ratio and light habitat is both
qualitative and quantitative. A higher proportion of sun leaves are
amphistomatous (\emph{39}) and the proportion of stomata on the upper,
adaxial surface increases with light (\emph{42}, \emph{43}). Resolving
why high light intensity favors amphistomatous dorsivental leaves is an
important first step toward understanding variation in stomatal density
ratio and leaf structure-function relationships more generally.

The overarching hypothesis is that leaves with greater stomatal density
ratio are more common in open, sunny habitats because they increase
photosynthesis most in those circumstances. An amphistomatous leaf
increases photosynthetic carbon gain compared to an otherwise identical
hypostomatous leaf by increasing conductance through the leaf
intercellular airspaces and boundary layers. We quantify this benefit as
the amphistomy advantage (\aax{}), the log-response ratio of
photosynthesis in an amphistomatous leaf compared to an otherwise
identical pseudohypostomatous leaf (\emph{20}, \emph{46}). Why would
\aax{} be greater in sun than shade? We consider three nonmutually
exclusive hypotheses that we classify as `acclimation', `plasticity',
and `constitutive' (Fig.~\ref{fig-hypotheses}).

@@

\emph{Plasticity hypothesis}: Individuals of the same genotype often
develop dramatically different leaves in sun and shade conditions
(\emph{47}). Plastic responses are likely adaptations to optimize
photosynthesis at different light intensities in variable environments
(\emph{48}). Plastic changes in leaf anatomy and biochemistry could
modulate \aax{} as a byproduct. Thicker or less porous leaves, both of
which are associated with high leaf mass per area (LMA), will have lower
\gias; leaves with increased total stomatal density and photosynthetic
capacity have greater potential CO\(_2\) supply and demand. Under the
plasticity hypothesis, we predict that \(\aasun{} > \aashade{}\) for all
populations and light intensities. \(\aasun{}\) is the \aax{} measured
on sun leaves; \(\aashade{}\) is the \aax{} measured on shade leaves.

\emph{Constitutive hypothesis}: In environments that are relatively
constant or where environmental change cannot be anticipated by a
reliable cue, natural selection will favor constitutive expression of
optimal phenotypes. We therefore predict genotypes from more sunny, open
habitats will have consistently greater \aax{} under all measurement and
growth light intensities. For herbaceous plants, light intensity is
largely a function of the tree canopy (\emph{49}). Herbs growing in the
open will regularly experience high light intensity; herbs growing under
a forest canopy will often experience low light intensity.

@@

We tested these hypotheses by comparing \aax{} among amphistomatous wild
tomato species (Table~\ref{tbl-populations}; Figure~\ref{fig-sr-rxnorm};
(\emph{50})) from different native light habitats, grown under simulated
sun and shade light treatments, and measured under contrasting light
intensities (low and high). We measured \aax{} on 572 individual plants
from 29 populations (average of 9.86 replicates per light treatment)
using a recently developed method (\emph{46}). With this method, we
directly compare the photosynthetic rate of an untreated amphistomatous
leaf to that of the same leaf with gas exchange blocked through the
adaxial (upper) surface by transparent plastic, which we refer to as
`pseudohypostomy'. To compare amphi- and pseudohypostomatous leaves at
identical whole-leaf stomatal conductance (\gsw), we measure \(A\) over
a range of \gsw, inducing stomatal opening and closure by modulating
humidity (see supplementary text for further details). We estimated
`amphistomy advantage' (\aax) \emph{sensu} (\emph{20}), but with
modifications previously described in (\emph{46}) and here
(Supplementary Materials). The native light intensity was represented by
plant area index (PAI \(\unit{\meter\squared\per\meter\squared}\)),
estimated using a global gridded data set derived from the Global
Ecosystem Dynamics Investigation {[}GEDI; (\emph{51}){]} and
georeferenced accession collection information from the Tomato Genetics
Resource Center (Table~\ref{tbl-populations}). The growth light
intensities were \ppfdequals{761} (sun treatment) and \ppfdqty{115}
(shade treatment) while all other environment conditions were nearly
identical (see Materials and Methods). The high and low measurement
light intensities were \ppfdequals{2000} (97.8:2.24 red:blue) and
\ppfdequals{150} (87.0:13.0 red:blue), respectively.

@@

The pattern of \aax{} across wild tomatoes strongly supports the
plasticity hypothesis, contradicts the acclimation hypothesis, and
provides modest support for the constitutive hypothesis. Plastic changes
in leaf thickness and/or packing density, summarized by the bulk leaf
mass per area (LMA), may mediate the effect of growth light intensity on
\aax{}. LMA increased in sun grown plants in all populations by an
average of 123\% {[}95\% CI: 42.9 to 256\%{]}, qualitatively similar to
plastic responses in many species (\emph{47}). While LMA is weakly,
albeit significantly (Table~\ref{tbl-fit_aa2}), associated with
individual-level \aax{} (Fig.~\ref{fig-lma_aa}), the effect of growth
light intensity on \aax{} is still predictive based on model comparison
using information criteria (Table~\ref{tbl-aa_loo1}). However, the
direct effect of the sun treatment on \aax{} was weaker when LMA was
included in the model (Table~\ref{tbl-fit_aa1} vs.
Table~\ref{tbl-fit_aa2}), suggestive of a mediating role. Many
anatomical traits underlie LMA (\emph{52}) and future research will be
needed to identify which particular traits, such as leaf thickness or
mesophyll porosity, are responsible for mediating \aax{}. The fact that
the \aax{} of sun plants was greater under low measurement light
intensity supports a long-standing hypothesis that resistance to
CO\(_2\) diffusion is greater in the upper than lower portions of the
leaf interior (\emph{53}). At low light intensity, photosynthesis is
weighted toward the upper palisade where most light is intercepted. If
resistance to CO\(_2\) diffusion is high in the upper palisade,
amphistomy may, unexpectedly, be particularly beneficial for sun leaves
experiencing intermittent shade or cloud cover. Our study is limited in
testing this because we could not directly measure the stomatal
conductance ratio and intercellular resistance on each surface. Future
experiments measuring \aax{} with a dual sided chamber (\emph{31},
\emph{33}) can overcome these limitations.

@@

@@

\bibitem[\citeproctext]{ref-mcadam_stomata_2021}
\CSLLeftMargin{2. }%
\CSLRightInline{S. A. M. McAdam, J. G. Duckett, F. C. Sussmilch, S.
Pressel, K. S. Renzaglia, R. Hedrich, T. J. Brodribb, A. Merced,
\href{https://doi.org/10.1002/ajb2.1619}{Stomata: The holey grail of
plant evolution}. \emph{American Journal of Botany} \textbf{108},
366--371 (2021).}

\bibitem[\citeproctext]{ref-clark_origin_2022}
\CSLLeftMargin{3. }%
\CSLRightInline{J. W. Clark, B. J. Harris, A. J. Hetherington, N.
Hurtado-Castano, R. A. Brench, S. Casson, T. A. Williams, J. E. Gray, A.
M. Hetherington, \href{https://doi.org/10.1016/j.cub.2022.04.040}{The
origin and evolution of stomata}. \emph{Current Biology} \textbf{32},
R539--R553 (2022).}

@@

\bibitem[\citeproctext]{ref-de_boer_optimal_2016}
\CSLLeftMargin{5. }%
\CSLRightInline{H. J. de Boer, C. A. Price, F. Wagner‐Cremer, S. C.
Dekker, P. J. Franks, E. J. Veneklaas,
\href{https://doi.org/10.1111/nph.13929}{Optimal allocation of leaf
epidermal area for gas exchange}. \emph{New Phytologist} \textbf{210},
1219--1228 (2016).}

@@

\bibitem[\citeproctext]{ref-franks_new_2014}
\CSLLeftMargin{9. }%
\CSLRightInline{P. J. Franks, D. L. Royer, D. J. Beerling, P. K. Van de
Water, D. J. Cantrill, M. M. Barbour, J. A. Berry,
\href{https://doi.org/10.1002/2014GL060457}{New constraints on
atmospheric {CO}\(_{\textrm{2}}\) concentration for the {Phanerozoic}}.
\emph{Geophysical Research Letters} \textbf{41}, 4685--4694 (2014).}

@@

\bibitem[\citeproctext]{ref-the_cenozoic_co_proxy_integration_project_cencopip_consortium_toward_2023}
\CSLLeftMargin{11. }%
\CSLRightInline{The Cenozoic CO Proxy Integration Project (CenCOPIP)
Consortium*†, B. Hönisch, D. L. Royer, D. O. Breecker, P. J. Polissar,
G. J. Bowen, M. J. Henehan, Y. Cui, M. Steinthorsdottir, J. C. McElwain,
M. J. Kohn, A. Pearson, S. R. Phelps, K. T. Uno, A. Ridgwell, E.
Anagnostou, J. Austermann, M. P. S. Badger, R. S. Barclay, P. K. Bijl,
T. B. Chalk, C. R. Scotese, E. De La Vega, R. M. DeConto, K. A. Dyez, V.
Ferrini, P. J. Franks, C. F. Giulivi, M. Gutjahr, D. T. Harper, L. L.
Haynes, M. Huber, K. E. Snell, B. A. Keisling, W. Konrad, T. K.
Lowenstein, A. Malinverno, M. Guillermic, L. M. Mejía, J. N. Milligan,
J. J. Morton, L. Nordt, R. Whiteford, A. Roth-Nebelsick, J. K. C.
Rugenstein, M. F. Schaller, N. D. Sheldon, S. Sosdian, E. B. Wilkes, C.
R. Witkowski, Y. G. Zhang, L. Anderson, D. J. Beerling, C. Bolton, T. E.
Cerling, J. M. Cotton, J. Da, D. D. Ekart, G. L. Foster, D. R.
Greenwood, E. G. Hyland, E. A. Jagniecki, J. P. Jasper, J. B. Kowalczyk,
L. Kunzmann, W. M. Kürschner, C. E. Lawrence, C. H. Lear, M. A.
Martínez-Botí, D. P. Maxbauer, P. Montagna, B. D. A. Naafs, J. W. B.
Rae, M. Raitzsch, G. J. Retallack, S. J. Ring, O. Seki, J. Sepúlveda, A.
Sinha, T. F. Tesfamichael, A. Tripati, J. Van Der Burgh, J. Yu, J. C.
Zachos, L. Zhang, \href{https://doi.org/10.1126/science.adi5177}{Toward
a {Cenozoic} history of atmospheric {CO}\(_{\textrm{2}}\)}.
\emph{Science} \textbf{382}, eadi5177 (2023).}

@@

\bibitem[\citeproctext]{ref-liang_stomatal_2023}
\CSLLeftMargin{13. }%
\CSLRightInline{X. Liang, D. Wang, Q. Ye, J. Zhang, M. Liu, H. Liu, K.
Yu, Y. Wang, E. Hou, B. Zhong, L. Xu, T. Lv, S. Peng, H. Lu, P. Sicard,
A. Anav, D. S. Ellsworth,
\href{https://doi.org/10.1038/s41467-023-37934-7}{Stomatal responses of
terrestrial plants to global change}. \emph{Nature Communications}
\textbf{14}, 2188 (2023).}

@@

\bibitem[\citeproctext]{ref-lang_century-long_2024}
\CSLLeftMargin{15. }%
\CSLRightInline{P. L. M. Lang, J. M. Erberich, L. Lopez, C. L. Weiß, G.
Amador, H. F. Fung, S. M. Latorre, J. R. Lasky, H. A. Burbano, M.
Expósito-Alonso, D. C. Bergmann,
\href{https://doi.org/10.1038/s41559-024-02481-x}{Century-long timelines
of herbarium genomes predict plant stomatal response to climate change}.
\emph{Nature Ecology \& Evolution} \textbf{8}, 1641--1653 (2024).}

>>>>

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\KOMAoption{captions}{tableheading}
\usepackage{hyperref}
\addtokomafont{disposition}{\rmfamily}
\usepackage{siunitx}
\newcommand{\aahigh}{\mathrm{AA}_{2000}}
\newcommand{\aalow}{\mathrm{AA}_{150}}
\newcommand{\aashade}{\mathrm{AA}_{\text{shade}}}
\newcommand{\aasun}{\mathrm{AA}_{\text{sun}}}
\newcommand{\aax}{$\mathrm{AA}$}
\newcommand{\aay}{\mathrm{AA}}
\newcommand{\Aamphi}{$A_{\mathrm{amphi}}$}
\newcommand{\Ahypo}{$A_{\mathrm{hypo}}$}
\newcommand{\agcurve}{$A \textendash g_\text{sw}$}
\newcommand{\aqcurve}{$A \textendash Q$}
\newcommand{\ca}{$C_\mathrm{a}$}
\newcommand{\caequals}[1]{$C_\mathrm{a} = \qty{#1}{\micro\mol\raiseto{-1}\mol}$}
\newcommand{\cabetween}[2]{#1 < $C_\mathrm{a} < \qty{#2}{\micro\mol\raiseto{-1}\mol}$}
\newcommand{\gias}{$g_\text{ias}$}
\newcommand{\gmax}{$g_\text{max}$}
\newcommand{\gmaxratio}{$g_\text{max,ratio}$}
\newcommand{\gmaxshade}{$g_\text{max,shade}$}
\newcommand{\gmaxsun}{$g_\text{max,sun}$}
\newcommand{\gsw}{$g_\text{sw}$}
\newcommand{\loggsw}{$\log(g_\text{sw})$}
\newcommand{\logA}{$\log(A)$}
\newcommand{\ppfd}{$\mathrm{PPFD}$}
\newcommand{\ppfdequals}[1]{$\mathrm{PPFD} = \qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
\newcommand{\ppfdqty}[1]{$\qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
\newcommand{\rh}{$\mathrm{RH}$}
\newcommand{\rhequals}[1]{$\mathrm{RH} = #1\%$}
\newcommand{\tleafequals}[1]{$T_\mathrm{leaf} = \qty{#1}{\degreeCelsius}$}
\usepackage{setspace}
\usepackage[modulo]{lineno}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

@@

\begin{center}
$^1$School of Life Sciences, University of Hawaiʻi; Mānoa, Hawaiʻi, USA\\
$^2$Department of Botany, University of Wisconsin, Madison; Wisconsin, USA\\
\medskip
*Corresponding author: cdmuir@wisc.edu\\
\medskip
ORCID: Christopher D. Muir — \href{https://orcid.org/0000-0003-2555-3878}{0000-0003-2555-3878}\\
\medskip
Classification: Biological Sciences, Ecology\\
\medskip
Keywords: amphistomy, photosynthesis, \textit{Solanum}, stomata
\end{center}

\newpage
\begin{linenumbers}
\begin{doublespace}

@@

\subsection{Significance statement}\label{significance-statement}

Stomata, the ``mouths'' of plants, regulate CO\(_2\) available for
photosynthesis. They are ubiquitous on leaves and essential to
understanding why vascular plants transformed Earth and how our crops
grow. Yet the adaptive significance of why many leaves only have stomata
on their lower surface is a mystery. Many assume that this limits water
loss, but we show the opposite is true. By manipulating which stomata a
leaf can use, we show that leaves with stomata on both surfaces gain
more CO\(_2\) for the same amount of water lost, increasing water use
efficiency. Furthermore, species from open canopies grown under high
light benefit the most, which may explain why leaves with stomata on
both sides are most prevalent in these habitats.

@@

Stomata are microscopic pores on the surfaces of leaves and other
photosynthetic organs formed by a pair of guard cells. They are
essential for balancing carbon gained per unit water lost and were an
essential innovation in vascular plants, permitting them to grow tall on
land by enabling access to CO\(_2\) for photosynthetic carbon
assimilation while preventing hydraulic failure in variable environments
(1--3). Optimal stomatal function depends on both dynamic changes in
aperture on the scale of minutes to hours, as well as static anatomy
determined by developmental plasticity and constitutive genetic
differences (4--6). Understanding how stomata respond to environmental
change over daily, developmental, and evolutionary time is important for
studying adaptation (1, 7, 8), inferring paleoclimate from fossil
cuticles (9--11), predicting responses to climate change (12--15), and
improving crops (16). Stomatal function contributes to global carbon and
water cycles (17) and therefore to predicting future climate (18).

Despite extensive theoretical and empirical progress understanding
stomata function and anatomy from molecular to ecosystem levels, the
adaptive significance of amphistomatous leaves remains an important
unsolved problem in leaf structure-function relationships (19--26).
Amphistomatous leaves develop abaxial and adaxial stomata whose aperture
can be independently regulated (27--31) to control gas exchange through
each surface. All else being equal, simultaneous gas exchange through
stomata on both surfaces increases CO\(_2\) supply to chloroplasts by
providing a second parallel pathway through leaf intercellular
airspaces, enhancing photosynthesis (20, 32). The extent to which
amphistomy increases CO\(_2\) supply depends on resistance to diffusion
in intercellular airspaces. This resistance can be low in thin, porous,
amphistomatous leaves (28, 33), but may be more substantial in thick,
dense, hypostomatous leaves (34). We refer to the intercellular airspace
conductance (\gias), the inverse of resistance. Amphistomatous leaves
also lose more water through evaporation because of a second boundary
layer conductance (35), but the additional carbon gain should be enough
to offset this cost in most realistic scenarios (36).

The paradoxical fact is that, despite the photosynthetic benefit, most
leaves are not amphistomatous. Many vertically oriented and/or
isobilateral leaves are amphistomatous (25). But among dorsiventral
leaves, it is primarily herbaceous plants in open, high light habitats
that tend to have amphistomatous leaves (22, 39--44). Most other leaves,
except those from aquatic habitats, are hypostomatous, producing stomata
only on the lower, abaxial surface. Even resupinate leaves develop
stomata on the lower, albeit adaxial surface (45), suggesting that leaf
orientation (lower vs.~upper) rather than leaf polarity (abaxial
vs.~adaxial) is causal. Stomatal density ratio, defined as the ratio of
upper to total stomatal density, is a quantitative metric of stomatal
distribution. The covariation between stomatal density ratio and light
habitat is both qualitative and quantitative. A higher proportion of sun
leaves are amphistomatous (39) and the proportion of stomata on the
upper, adaxial surface increases with light (42, 43). Resolving why high
light intensity favors amphistomatous dorsivental leaves is an important
first step toward understanding variation in stomatal density ratio and
leaf structure-function relationships more generally.

The overarching hypothesis is that leaves with greater stomatal density
ratio are more common in open, sunny habitats because they increase
photosynthesis most in those circumstances. An amphistomatous leaf
increases photosynthetic carbon gain compared to an otherwise identical
hypostomatous leaf by increasing conductance through the leaf
intercellular airspaces and boundary layers. We quantify this benefit as
the amphistomy advantage (\aax{}), the log-response ratio of
photosynthesis in an amphistomatous leaf compared to an otherwise
identical pseudohypostomatous leaf (20, 46). Why would \aax{} be greater
in sun than shade? We consider three nonmutually exclusive hypotheses
that we classify as `acclimation', `plasticity', and `constitutive'
(Fig.~\ref{fig-hypotheses}).

@@

\emph{Plasticity hypothesis}: Individuals of the same genotype often
develop dramatically different leaves in sun and shade conditions (47).
Plastic responses are likely adaptations to optimize photosynthesis at
different light intensities in variable environments (48). Plastic
changes in leaf anatomy and biochemistry could modulate \aax{} as a
byproduct. Thicker or less porous leaves, both of which are associated
with high leaf mass per area (LMA), will have lower \gias; leaves with
increased total stomatal density and photosynthetic capacity have
greater potential CO\(_2\) supply and demand. Under the plasticity
hypothesis, we predict that \(\aasun{} > \aashade{}\) for all
populations and light intensities. \(\aasun{}\) is the \aax{} measured
on sun leaves; \(\aashade{}\) is the \aax{} measured on shade leaves.

\emph{Constitutive hypothesis}: In environments that are relatively
constant or where environmental change cannot be anticipated by a
reliable cue, natural selection will favor constitutive expression of
optimal phenotypes. We therefore predict genotypes from more sunny, open
habitats will have consistently greater \aax{} under all measurement and
growth light intensities. For herbaceous plants, light intensity is
largely a function of the tree canopy (49). Herbs growing in the open
will regularly experience high light intensity; herbs growing under a
forest canopy will often experience low light intensity.

@@

We tested these hypotheses by comparing \aax{} among amphistomatous wild
tomato species (Table~\ref{tbl-populations}; Figure~\ref{fig-sr-rxnorm};
(50)) from different native light habitats, grown under simulated sun
and shade light treatments, and measured under contrasting light
intensities (low and high). We measured \aax{} on 572 individual plants
from 29 populations (average of 9.86 replicates per light treatment)
using a recently developed method (46). With this method, we directly
compare the photosynthetic rate of an untreated amphistomatous leaf to
that of the same leaf with gas exchange blocked through the adaxial
(upper) surface by transparent plastic, which we refer to as
`pseudohypostomy'. To compare amphi- and pseudohypostomatous leaves at
identical whole-leaf stomatal conductance (\gsw), we measure \(A\) over
a range of \gsw, inducing stomatal opening and closure by modulating
humidity (see supplementary text for further details). We estimated
`amphistomy advantage' (\aax) \emph{sensu} (20), but with modifications
previously described in (46) and here (Supplementary Materials). The
native light intensity was represented by plant area index (PAI
\(\unit{\meter\squared\per\meter\squared}\)), estimated using a global
gridded data set derived from the Global Ecosystem Dynamics
Investigation {[}GEDI; (51){]} and georeferenced accession collection
information from the Tomato Genetics Resource Center
(Table~\ref{tbl-populations}). The growth light intensities were
\ppfdequals{761} (sun treatment) and \ppfdqty{115} (shade treatment)
while all other environment conditions were nearly identical (see
Materials and Methods). The high and low measurement light intensities
were \ppfdequals{2000} (97.8:2.24 red:blue) and \ppfdequals{150}
(87.0:13.0 red:blue), respectively.

@@

The pattern of \aax{} across wild tomatoes strongly supports the
plasticity hypothesis, contradicts the acclimation hypothesis, and
provides modest support for the constitutive hypothesis. Plastic changes
in leaf thickness and/or packing density, summarized by the bulk leaf
mass per area (LMA), may mediate the effect of growth light intensity on
\aax{}. LMA increased in sun grown plants in all populations by an
average of 123\% {[}95\% CI: 42.9 to 256\%{]}, qualitatively similar to
plastic responses in many species (47). While LMA is weakly, albeit
significantly (Table~\ref{tbl-fit_aa2}), associated with
individual-level \aax{} (Fig.~\ref{fig-lma_aa}), the effect of growth
light intensity on \aax{} is still predictive based on model comparison
using information criteria (Table~\ref{tbl-aa_loo1}). However, the
direct effect of the sun treatment on \aax{} was weaker when LMA was
included in the model (Table~\ref{tbl-fit_aa1} vs.
Table~\ref{tbl-fit_aa2}), suggestive of a mediating role. Many
anatomical traits underlie LMA (52) and future research will be needed
to identify which particular traits, such as leaf thickness or mesophyll
porosity, are responsible for mediating \aax{}. The fact that the \aax{}
of sun plants was greater under low measurement light intensity supports
a long-standing hypothesis that resistance to CO\(_2\) diffusion is
greater in the upper than lower portions of the leaf interior (53). At
low light intensity, photosynthesis is weighted toward the upper
palisade where most light is intercepted. If resistance to CO\(_2\)
diffusion is high in the upper palisade, amphistomy may, unexpectedly,
be particularly beneficial for sun leaves experiencing intermittent
shade or cloud cover. Our study is limited in testing this because we
could not directly measure the stomatal conductance ratio and
intercellular resistance on each surface. Future experiments measuring
\aax{} with a dual sided chamber (31, 33) can overcome these
limitations.

@@

\end{doublespace}
\end{linenumbers}

@@

\bibitem[\citeproctext]{ref-mcadam_stomata_2021}
\CSLLeftMargin{2. }%
\CSLRightInline{S. A. M. McAdam, \emph{et al.},
\href{https://doi.org/10.1002/ajb2.1619}{Stomata: The holey grail of
plant evolution}. \emph{American Journal of Botany} \textbf{108},
366--371 (2021).}

\bibitem[\citeproctext]{ref-clark_origin_2022}
\CSLLeftMargin{3. }%
\CSLRightInline{J. W. Clark, \emph{et al.},
\href{https://doi.org/10.1016/j.cub.2022.04.040}{The origin and
evolution of stomata}. \emph{Current Biology} \textbf{32}, R539--R553
(2022).}

@@

\bibitem[\citeproctext]{ref-de_boer_optimal_2016}
\CSLLeftMargin{5. }%
\CSLRightInline{H. J. de Boer, \emph{et al.},
\href{https://doi.org/10.1111/nph.13929}{Optimal allocation of leaf
epidermal area for gas exchange}. \emph{New Phytologist} \textbf{210},
1219--1228 (2016).}

@@

\bibitem[\citeproctext]{ref-franks_new_2014}
\CSLLeftMargin{9. }%
\CSLRightInline{P. J. Franks, \emph{et al.},
\href{https://doi.org/10.1002/2014GL060457}{New constraints on
atmospheric {CO}\(_{\textrm{2}}\) concentration for the {Phanerozoic}}.
\emph{Geophysical Research Letters} \textbf{41}, 4685--4694 (2014).}

@@

\bibitem[\citeproctext]{ref-the_cenozoic_co_proxy_integration_project_cencopip_consortium_toward_2023}
\CSLLeftMargin{11. }%
\CSLRightInline{The Cenozoic CO Proxy Integration Project (CenCOPIP)
Consortium*†, \emph{et al.},
\href{https://doi.org/10.1126/science.adi5177}{Toward a {Cenozoic}
history of atmospheric {CO}\(_{\textrm{2}}\)}. \emph{Science}
\textbf{382}, eadi5177 (2023).}

@@

\bibitem[\citeproctext]{ref-liang_stomatal_2023}
\CSLLeftMargin{13. }%
\CSLRightInline{X. Liang, \emph{et al.},
\href{https://doi.org/10.1038/s41467-023-37934-7}{Stomatal responses of
terrestrial plants to global change}. \emph{Nature Communications}
\textbf{14}, 2188 (2023).}

@@

\bibitem[\citeproctext]{ref-lang_century-long_2024}
\CSLLeftMargin{15. }%
\CSLRightInline{P. L. M. Lang, \emph{et al.},
\href{https://doi.org/10.1038/s41559-024-02481-x}{Century-long timelines
of herbarium genomes predict plant stomatal response to climate change}.
\emph{Nature Ecology \& Evolution} \textbf{8}, 1641--1653 (2024).}
